---
title: "Final Project"
author: "Thomas Simons and Joey Cerquera"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---


```{r setup, echo=FALSE}
library(knitr)
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
```


```{r}
library(tidyverse)
library(tree)
library(randomForest)
library(gbm)
library(ROCR)
library(e1071)
library(imager)   #load libraries 
library(plyr)
library(class)
library(rpart)
library(maptree)
library(ROCR)
```


```{r}
white_wine <- read.csv("~/UCSB HW R/PSTAT 131 Final/winequality-white.csv", sep = ";")
red_wine <- read.csv("~/UCSB HW R/PSTAT 131 Final/winequality-red.csv", sep = ";")
#import data 


white_wine <- mutate(white_wine, class = factor(ifelse(white_wine$quality > 5, "good", "bad"), levels = c("good", "bad")))

red_wine <- mutate(red_wine, class = factor(ifelse(red_wine$quality > 5, "good", "bad"), levels = c("good", "bad"))) 
#create new factor called class for qood or bad wine qulity 
```


1. Exploratory Analysis 

```{r}
par(mfrow=c(1, 2))

red_hist <- hist(red_wine$quality)  #histogram for red wine 
white_hist <- hist(white_wine$quality)  #histogram for white wine 
```


```{r}
#Outliers are observations with quality not equal to 5 or 6 

dim(filter(red_wine, quality >= 7))[1] #number of obs with quality greater than or equal to 7
dim(filter(red_wine, quality < 5))[1] #number of obs with quality less than 5 

dim(filter(white_wine, quality >= 7))[1] #number of obs with quality greater than or equal to 7
dim(filter(white_wine, quality < 5))[1] #number of obs with quality less than 5 

```
   
Quality     >6     <5

Obs Red     217    63
Obs White   1060   183

It is interesting that for both red and white wines there were more observations that had a quality greater than 6. This means that for both red and white wine. 

```{r}
red_mean <- colMeans(select(red_wine, -class))
red <- as.data.frame(red_mean)

means <- mutate(red, white_mean = colMeans(select(white_wine, -class)))
#create data frame with the mean values of both datasets 

print(means)
```

We can see that most of the variable have different mean values for white and red wine. 
The ones that have similar mean values are density, pH, alcohol and quality. 

```{r}
summary(white_wine)
summary(red_wine)
```

2. Data Mining/Analysis

```{r}
train_white <- sample(1:nrow(white_wine),(.5*nrow(white_wine)))
white_train <- white_wine[train_white,]
white_test <- white_wine[-train_white,]
```

```{r}
plot(white_train$quality, type="p")
#points(ifelse(white_train$class=="good",col="red",col="blue"))
```


```{r}
white_wine2 <- white_wine[,c(1:11,12)]
glm.fit = glm(quality ~.,data=white_wine2, family=binomial)
plot(glm.fit)

```




```{r}
set.seed(1)
tune.out=tune(svm,quality~.,data=white_train,kernel="linear",
ranges=list(cost=c(0.001, 0.01, 0.1,1,5)))
summary(tune.out)

```


```{r}
bestmod=tune.out$best.model
summary(bestmod)
```


Logistic Regression

```{r}



```


